Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-07-14 cover_letter/response_letter.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 1.0 (28/12/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
% with extensive modifications by Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% may not be compatible with rest of template
% \documentclass[journal=jpcbfk,manuscript=article]{achemso}

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

%\documentclass[11pt,a4paper,draft]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc) 
\usepackage{letterbib}
\usepackage{graphicx} % Required for including pictures
\usepackage{microtype} % Improves typography
\usepackage{fancyhdr}
\usepackage[T1]{fontenc} % Required for accented characters
\usepackage[sort&compress,numbers,super]{natbib}
\usepackage{caption}
\usepackage{newfloat}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{soul}
\usepackage[svgnames]{xcolor}
\usepackage{framed}
\usepackage{calligra}
\usepackage{fp}
\FPset\marginWidth{0.7}	% margin width (all sides) in inches
\usepackage[margin=\marginWidth in]{geometry}
\usepackage{marginfix}
\usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter

\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%more new commands and definitions
%\newcommand*{\rood}[1]{\color{red}{#1}}
\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\noteg}[1]{\textcolor{green}{[[#1]]}}		% notes

\newcommand*{\blauw}[1]{\color{blue}{#1}}
\newcommand*{\groen}[1]{\color{green}{#1}}
\newcommand{\hlc}[2][yellow]{{\sethlcolor{#1}\hl{#2}}}
\newcommand{\scripty}[1]{\ensuremath{\mathcalligra{#1}}}

% code for custom comments
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{mathtools}

% markers (black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, olive, orange, pink, purple, red, teal, violet, white, yellow, more: https://en.wikibooks.org/wiki/LaTeX/Colors)
\newcommand{\marker}[3]{
  \tikz[baseline=(X.base)]{
    \node [fill=#1!40,rounded corners] (X) {#2:};
  }
  {\color{#1!80!black}#3}
}
\newcommand{\khb}[1]{\marker{teal}{KHB}{#1}}  % Kathy Breen
\newcommand{\replyhead}{\textbf{Author Reply: }}

\newfloat{figure}{h}{lot}
\floatname{figure}{Figure}
\newfloat{sfigure}{h}{lot}
\floatname{sfigure}{Figure}
\newfloat{lfigure}{h}{lot}
\floatname{lfigure}{Figure}

\definecolor{shadecolor}{named}{LightGray}

\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\DeclareFontShape{T1}{calligra}{m}{n}{<->s*[2.2]callig15}{}

\renewcommand{\thesfigure}{S\arabic{sfigure}}
\renewcommand{\thelfigure}{\Alph{lfigure}}

%-----------------------------------------------------------------------------
% HEADER/FOOTER INFO
%-----------------------------------------------------------------------------
%\pagestyle{fancy}
\fancyhf{}
%\cfoot{\textit{telephone} 217-333-1441 $\bullet$ \textit{fax} 217-333-2736 $\bullet$ \textit{email} matse@illinois.edu}
\cfoot{\thepage}

%----------------------------------------------------------------------------------------
%	SENDER INFORMATION
%----------------------------------------------------------------------------------------

\def\Who{Andrew L. Ferguson} % Your name
\def\What{, PhD} % Your title
\def\Where{Pritzker School of Molecular Engineering\\University of Chicago} % Your department/institution
\def\department{Pritzker School of Molecular Engineering}
\def\institution{University of Chicago}
\def\Address{5640 South Ellis Avenue} % Your address
\def\CityZip{Chicago, IL 60637} % Your city, zip code, country, etc
\def\Email{e:~\url{andrewferguson@uchicago.edu}} % Your email address
\def\TEL{p:~(773)~702-5950} % Your phone number
\def\URL{w:~\url{www.ferglab.com}} % Your URL
\def\titleone{Associate Professor\\ Deputy Dean for Equity, Diversity, and Inclusion}

%----------------------------------------------------------------------------------------
%	HEADER AND FROM ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing institution logo indent based on margin width
\FPsub\logoIndent{7.55}{\tmp}

\address{
\includegraphics[width=3.5in]{UChicago_PME_logo.png} % Include the logo of your institution
%\hspace{\logoIndent in} % Position of the institution logo, increase to move left, decrease to move right
\hspace{2.8in} % Position of the institution logo, increase to move left, decrease to move right
\vskip -0.82in~\\ % Position of the text in relation to the institution logo, increase to move down, decrease to move up
\Large\hspace{0.75in} \hfill ~\\[0.05in] % First line of institution name, adjust hspace if your logo is wide
\hspace{0.75in} \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
\makebox[0ex][r]{\bf \Who \What }\hspace{0.65in} % Print your name and title with a little whitespace to the right
~\\[-0.11in] % Reduce the whitespace above the horizontal rule
\hspace{0in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
\hspace{\fill}\parbox[t]{4in}{ % Create a box for your details underneath the horizontal rule on the right
\footnotesize % Use a smaller font size for the details
%\Who \\  % Your name, all text after this will be italicized
%\\
\titleone
%\titletwo\\
%\titlethree\\
%\titlefour\\
\vskip 0.05in
\department\\ % Your department
\institution\\ % Your institution
\Address\\ % Your address
\CityZip
\vskip 0.05in
\TEL\\ % Your phone number
\Email\\ % Your email address
\URL\\ % Your URL
}
\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
\vspace{-1in} % Move the letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	TO ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\FPmul\tmp{2}{\marginWidth}	% computing date indent based on margin width
\FPsub\dateIndent{4.15}{\tmp}

\def\opening#1{\thispagestyle{empty}
{\centering\fromaddress \vspace{1.1in} \\ % Print the header and from address here, add whitespace to move date down
%
\hspace*{4.15 in}\today\hspace*{\fill}\par
} % Print today's date, remove \today to not display it
{\raggedright \toname \\ \toaddress \par} % Print the to name and address
\vspace{0.05in} % White space after the to address
\noindent #1 % Print the opening line
% Uncomment the 4 lines below to print a footnote with custom text
%\def\thefootnote{}
%\def\footnoterule{\hrule}
%\footnotetext{\hspace*{0.18\textwidth}{Telephone 217-333-1441 $\bullet$ Fax 217-333-2736 $\bullet$ email matse@illinois.edu\footnotesize \hspace{3cm} \thepage}}
%\def\thefootnote{\arabic{footnote}}
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who \What} % The signature is a combination of your name and title

\long\def\closing#1{
\vspace{0.1in} % Some whitespace after the letter content and before the signature
\noindent % Stop paragraph indentation
%\hspace*{\longindentation} % Move the signature right
%\parbox{\indentedwidth}{\raggedright
#1 % Print the signature text
\vskip 0.0in
\includegraphics[width=0.2\textwidth]{signature_720}
%\vskip 0.65in % Whitespace between the signature text and your name
\\
\noindent
\fromsig
\vskip -0.05in
\titleone\\
\department\\
\institution\\
\vskip -0.1in
\noindent cc: M.S.~Jones, B.~Ashwood, A.~Tokmakoff
}

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}
{
Prof.\ Hanadi Sleiman \\
Associate Editor, \textit{The Journal of the American Chemical Society}
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Prof.\ Sleiman}

\noteg{update below}

Thank you for your 23-Jul-2020 message regarding our article “Molecular Latent Space Simulators” assigned ID SC-EDG-07-2020-003635. We thank you for handling our submission and the anonymous reviewers for their careful reading and critical assessment of our work.

We reproduce below the full text of the reviewer comments in blue along with our responses, reproduction of the changes made to our manuscript, and the locations of these edits. The modifications within our revised submission in response to the reviewer comments are indicated in red text.

In light of these changes, we hope that our revised submission is suitable for publication in Chemical Science.

\closing{Yours Sincerely}

\end{letter}

%----------------------------------------------------------------------------------------

\clearpage
\newpage

Brennan comments:
\begin{enumerate}
    \item Get rid of "elevated" fraying response
    \item Rev 1, either view to show
    \item Figure out if the conc correction makes sense -- how much does it actually improve things?
    \item Kd hybridization does not actually increase with temp
\end{enumerate}

\begin{shaded}
\textbf{Response to Reviewer \# 1}
\end{shaded}

\textcolor{blue}{In the manuscript “Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy”, Jones et al. provide a well-designed and compelling study of DNA hybridization kinetics and dynamics that succinctly encompasses both experiment and theory. They demonstrate that the coarse-grained model of DNA, 3SPN.2, can not only well-reproduce the relative kinetic rates of sequence-dependent hybridization but also can reveal critical insights into their physical underpinnings. These insights are driven by the use of state-of-the-art applications of machine learning to Markov state model construction and well-validated with temperature jump infrared spectroscopy. In particular, the demonstration of how the placement of two G:C base pairs within the sequence can dramatically affect the energy landscape is truly impactful for the design of DNA sequences with desired properties. We believe this work is well suited for JACS publication after the Authors address a few major and minor points.}

\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased they find it significant to JACS readers.

\textcolor{blue}{Major point.
1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D. While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ‘representative’ structures shown in Figures 2A, C). A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern. This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations. At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
}

\textbf{Author Reply:}   We acknowledge that this information in crucial to the core conclusions of the manuscript and that some information regarding state definitions and representative structures was unclear.

When defining macrostates, we used both visual inspection of configurations and plots along a variety of structural metrics. In order to demonstrate that configuration in a given macrostates are well-defined by the representations in figure 2, we show the distribution of several structural metrics for each macrostate and sequence combination. These metrics are a collection of mean intermolecular distance that we have found work well to differentiate between states.

During the analysis process, we have found that two distance features in particular allow for interpretable visualization of each of the seven macrostate. Specifically, these features correspond to the reciprocal mean distance between 2-base-shifted base pairs in the 5\prime and 3\prime directions. We have visualized these features below, and use the PyEMMA free energy function to plot the distribution of each macrostate/sequence on the low dimensional feature representation. Each color map corresponds to a different macrostate according to the legend, and lighter regions correspond to higher probability at equilibrium (lower free energy).

We agree with the reviewer that it is important readers to understand the physical definitions and distributions of each state, and we have added this figure to the supplemental information.

\noteg{state distribution figure}

With regard the representative states shown in figure 2A, these were chosen from microstates with the highest probability of being assigned to the macrostates they represent. Accordingly, they should correspond to structures  that are representative of that macrostate as whole. As the reviewer points out, it is unclear how these selected microstates were defined and we include the following addition to section 3.2 to make this point explicit:

``We present these seven macrostates in Fig. 2a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. \rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole''}

\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering. I'be tried to address both aspect above, but I think}

\textcolor{blue}{1)      It is unclear how five datasets were selected used for cross-validation in the SRVs (P. 10 L. 43). One would naively think a 50:50 split of a single trajectory would result in only one dataset. A bit more detail could be provided here.}

\textbf{Author Reply:}   Five-fold cross-validation was conducted by randomly partitioning 40 independent trajectories into two sets of 20. In this way, all data is included in each fold but with distinct test/train sets. This approach prevents rare events from being left out of small validation sets and was used in our previous study, where one long trajectory was broken up into 100 \citep{Sidky2019High-ResolutionVAMPnets} smaller trajectories for training. We have  clarified this in the text by adding the following sentence to section 2.1:

We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits \rood{such that random combinations of 20 trajectories were partitioned into each set}.\citep{Sidky2019High-ResolutionVAMPnets}

\textcolor{blue}{2)      For the validation of embedding dimensions (P. 11 L. 3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Figure S1. Could a few additional dimensions be provided?}

\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence. As suggested by the reviewer, we added two additional modes to the cross-validation procedure showing this kink after five modes and have updated the AT-all subplot of Figure S1 as follows in Figure \ref{fig:AT-all_crossval}.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=100mm]{cover_letter/revision_figures/AT-all_added_modes.png}
        \caption{Cross-validation for the AT-all sequence with two additional modes to show the presence of a kink in the VAMP-2 score after the fifth mode.}
        \label{fig:AT-all_crossval}
	\end{center}
\end{figure}

\textcolor{blue}{3)      The values associated with the “slow” and “fast” rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}

\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 14. We agree that this sentence in unclear in its current form and have made the following change:

``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes on the 10-30 $\mu$s and 70-100 ns timescales, respectively''}

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Figure 1 has been updated to include units in s$^{-1}$. In accordance with suggestions by reviewer \#3 we have also displayed the slow responses on a y-log axis, and used the finite size effect correction mentioned by reviewer \#2 to improve the fit (Figure \ref{fig:conc_correction}). \noteg{Should we include the figure here instead? Thought it made more sense in the R2 section since we are addressing a more substantial change}

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all Figure S4 plots to include goodness-of-fit values determined by the $R^2$ values of log fits (Figure \ref{goodness-of-fit}). We have also re-plotted these on the log axis for easier comparisons between the simulated data and fits. Although all fits are strong, it does appear that GC-core may be slightly worse than the others sequences. As the reviewer suggests, this may be a contributing factor to the weaker experimental agreement for GC-core as well, however it is difficult for us to make that claim definitely given the data.  \noteg{Not sure if we really want to go into this or update the main text?}

### 2021-07-14 manuscript_marked_up.tex
"response letter figures"

%\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{{#1}}

@@

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA.\citep{Chen, Sidky2019High-ResolutionVAMPnets} SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2011MarkovValidation} As observed by Husic and Pande, the lag time cannot be treated as a hyperparameter to be optimized via the VAMP-2 score, but must be selected as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to high-resolution Markovian macrostate MSMs. We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.

@@

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex. This cutoff was selected as the distance beyond with the strands are effectively non-interacting and defines the dissociated state. We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction. We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.

@@

We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models. MSMs define the long-lived metastable macrostates of the system, their equilibrium occupancies, and the equilibrium transition probabilities between them. As such, they are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates. We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences. Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which four terminal A:T base pairs are unbound, and the fully dissociated state (D). We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium. By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F). In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates. The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model. The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hybridization/dehybridization pathways. Immediately, we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas at the other end of the spectrum GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior. In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM. These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates. The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization. Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization. The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways. The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes. The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization. We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.

>>>>

\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{{#1}}

@@

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA.\citep{Chen, Sidky2019High-ResolutionVAMPnets} SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2011MarkovValidation} As observed by Husic and Pande, the lag time cannot be treated as a hyperparameter to be optimized via the VAMP-2 score, but must be selected as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to high-resolution Markovian macrostate MSMs. We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits \rood{such that random combinations of 20 trajectories were partitioned into each set}. \citep{Sidky2019High-ResolutionVAMPnets} We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training requiring approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.

@@

To computationally mimic the T-jump process in our 3SPN.2 simulations, we conducted 1 $\mu$s simulations of an initially hybridized DNA duplex over a range of temperatures and monitored its structural relaxation. We performed 120 independent simulations for each DNA sequence at each temperature, and from these extracted computational estimates of $k_d^\mathrm{slow}$ and $k_d^\mathrm{fast}$ (\blauw{Fig.~S4}). First, we tracked the slow response corresponding to duplex dissociation in our simulations by compiling the distribution of times at which both of the central base pairs first separate to a distance of 1.3 nm starting from an initial fully hybridized duplex. This cutoff was selected as the distance beyond which the strands are effectively non-interacting and defines the dissociated state. We extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. Second, we tracked the fast response corresponding to terminal base pair fraying by compiling the distribution of times at which either of the terminal base pairs first separated to a distance of 1.3 nm, corresponding to a complete breakage of the WC interaction. We extracted our computational estimate of $k_d^\mathrm{fast}$ through a decaying exponential fit to the fraction of unfrayed sequences as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$.

@@

We now proceeded to construct Markov state models (MSMs) from 1 ms of aggregated simulation trajectories for each of the four sequences at their respective melting temperatures to generate sequence-dependent kinetic models. MSMs define the long-lived metastable macrostates of the system, their equilibrium occupancies, and the equilibrium transition probabilities between them. As such, they are extremely valuable in providing both a quantitative predictive model and a physically comprehensible mechanistic understanding of the long-time dynamical evolution of the system between an ensemble of metastable macrostates. We present in Fig.~\ref{fig:allseq_table} the inferred MSMs for each of the four 10-base DNA sequences. Across all four sequences we identify a totality of seven metastable macrostates corresponding to the fully hybridized state (H) in which all native base pairings are intact, four shifted states in which the strands are translated out-of-register by two or four bases in the 5$^\prime$ (5S2, 5S4) or 3$^\prime$ (3S2, 3S4) direction, a frayed state (F4) -- unique to GC-core -- in which four terminal A:T base pairs are unbound, and the fully dissociated state (D). We present these seven macrostates in Fig.~\ref{fig:allseq_table}a along with schematic and cartoon renderings of representative microstates contained within each of these macrostates. \rood{Molecular renderings were sampled from microstates with the highest probability of being clustered into their respective PCCA+ macrostates and should therefore reflect the macrostate population as whole.} In Fig.~\ref{fig:allseq_table}b we present the occupancy probabilities of each state at thermodynamic equilibrium. By virtue of the fact that each sequence is simulated at its corresponding melting temperature (T$_\mathrm{m}$), the probability of the dissociated state (D) is, by construction, approximately equal to the sum of the probabilities over the remaining six states (H, 5S2, 3S2, 5S4, 3S4, F). In Fig.~\ref{fig:allseq_table}c we present a visualization of the macrostate MSMs for each of the four sequences showing the connectivity between the identified macrostates. The macrostates are represented as orange circles in proportion to their equilibrium probabilities and the grey arrows indicate the probability of hopping from one macrostate to another under one time step of the kinetic model. The fluxes between the macrostates provide a wealth of high-level, interpretable information on the sequence-dependent metastable states and hybridization/dehybridization pathways. Immediately, we identify that the AT-all sequence possesses a rich and complex dynamical landscape comprising six metastable states whereas at the other end of the spectrum GC-mix exhibits far simpler two-state ``all-or-nothing'' behavior. In Fig.~\ref{fig:allseq_table}d we present the so-called implied time scales of each MSM. These time scales correspond to the relaxation times of the DNA dimer among its constituent metastable macrostates. The leading implied time scale for each system corresponds to the characteristic time scale for hybridization/dehybridization. Since each system is simulated at the same concentration and at its respective melting temperature, it is not surprising that the leading time scale is approximately equal for all four systems and corresponds to the characteristic time scale for hybridization/dehybridization. The spectrum of higher order time scales corresponds to increasingly quicker relaxations between the metastable macrostates within the kinetic model and resolve the interesting sequence-dependent differences in the hybridization/dehybridization kinetic pathways. The total number of implied time scales is typically one fewer than the number of metastable macrostates and the existence of large implied time scales is indicative of slowly relaxing kinetic processes. The dense spectrum of slow implied time scales for AT-all is indicative of its relatively complex kinetic landscape whereas that for the two-state GC-mix comprises only a single time scale corresponding to hybridization/dehybridization. We now proceed to analyze in detail the sequence-dependent thermodynamics, kinetic, and mechanisms exposed by the four MSMs.
